Clinical prediction using similarity cohort--based models: A systematic scoping review

### 2025-03-14 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.Rmd
put latex revisions into rmarkdown

We conducted all searches over 2021 July 19–22. We tentatively excluded results from Google Scholar because they were missing abstracts, and later agreed to discard these results due to the lack of reproducibility of the search. We organized the remaining results into a public Zotero collection with one subcollection for each database. We then imported the pooled results to Covidence, which merged some duplicate entries.

@@

These programs used varying terminology for common concepts, and no common term is in use for what we here refer to as \hl{similarity cohort--based modeling}; authors described their approaches as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020; @Liu2022; @Doborjeh2022], and "precision cohort" analysis [@Wang2019; @Tang2021; @Ng2021]. \hl{(Interestingly, none of the articles described their methods as ``explainable'' ML.)} We find uses of "targeted", "personalized", and "precision" generic and imprecise, while transductive inference is an established term for a broader set of methods in ML.
\hl{While \emph{local} naturally contrasts with \emph{global}, the overloaded term \emph{localized models} stands to cause more confusion than it resolves, so we use \emph{similarity cohort--based models} to refer to this specific approach.}

@@

This relatively small sample exhibited a wide range of \hl{designs in both the assembly of pre-processing, retrieval, reuse, and interpretation steps and the specific tools deployed at each}. Discussion of the \hl{diversity} of \hl{tools} has been summarized above and published in greater detail in previous reviews [@Choudhury2016; @Sharafoddini2017; @Parimbelli2018]\hl{. In this section, we focus on the diversity of assemblies. These depend on} three largely independent choices \hl{(Figure }\ref{fig:framework}\hl{})}: (1) how to measure the similarity between cases, (2) how to retrieve a cohort of cases similar to an index case, and (3) how to generate a prediction or other statistical insight for the index case from the similarity cohort.

@@

\hl{A SCBM implementation based on this framework would help investigators systematically compare predictions and other output obtained using different tools for cohort retrieval and predictive modeling, as well as tune the parameters governing these tools. It would also make like-kind comparisons of performance easier on different tasks using different data sets and expedite validation studies. This would address the core technical needs expressed in our sample (Section }\ref{identified-needs}\hl{)} as well as the limitations of our own comparison of reported performances (Section }\ref{evaluations}\hl{).
We also believe that this framework could be used to make the diversity of possible approaches more comprehensible to users, enabling them to most appropriately tailor these decisions to their tasks.}

@@

\hl{Most studies were premised on the potential predictive value of localized models (Section }\ref{rationales}\hl{), but not all experiments affirmed this premise (Section }\ref{evaluations}\hl{). Those that did mostly used DML to improve retrieval and sometimes iteratively optimized the metric and the model.
However, even those that showcased several such specializations }[@CampilloGimenez2013; @Ng2015; @Zhang2018; @Liu2022]\hl{ mostly used their own precursors or commonplace alternatives as comparators. Future work should compare competitive alternatives to assess their performance and other trade-offs.
This work would be facilitated by standardization and modularization using a framework like that outlined above.
For example, }@CampilloGimenez2013\hl{ and }@Liu2022\hl{ use a modular design to toggle multiple choices and obtain neatly nested comparisons.}

@@

The search will have been designed to recover studies of the kind reviewed by the review papers from which we obtained our “seed set”. To validate the final search design, we will determine how many of the papers in this seed set that are indexed by PubMed are actually recovered by our search. In most cases, the focus of a review paper is different from ours, so we will only perform this validation test on the seed set obtained from two review papers that are (a) closest in focus to ours and (b) use terminology associated with the two distinct sub-literatures relevant to our focus: Choudhury & Begum (2016), which focuses on case-based reasoning in medicine, and Sharafoddini, Dubin, & Lee (2017), which focuses on patient similarity–based prediction models on health data. The proportion of each PubMed-indexed seed set that is recovered from our PubMed search provides a rough and optimistic yet useful estimate of the proportion of the relevant literature that our full search strategy will recover.

Once we have finalized the search as a logical pattern, we will take the following steps to generate the sample/corpus of literature that will be the starting point for our selection process.

1. The logical pattern will be converted to a search string using the syntax appropriate to each database in our search strategy. These include PubMed (already done as part of the search design), Web of Science, Academic Search Premier/Elite, and Mathematical Reviews.
2. The search will be conducted on each database and the results organized into a Zotero collection, with one subcollection for each database.
3. Duplicate results will be identified and merged. (A result obtained from multiple databases should have only one Zotero entry but should be filed under the subcollection for each database in which it was found.)

@@

@@

>>>>

We conducted all searches over 2021 July 19–22. We tentatively excluded results from Google Scholar because they were missing abstracts, and later agreed to discard these results due to the lack of reproducibility of the search.

@@

@@

This relatively small sample exhibited a wide range of \hl{designs in both the assembly of pre-processing, retrieval, reuse, and interpretation steps and the specific tools deployed at each}. Discussion of the \hl{diversity} of \hl{tools} has been summarized above and published in greater detail in previous reviews [@Choudhury2016; @Sharafoddini2017; @Parimbelli2018]\hl{. In this section, we focus on the diversity of assemblies. These depend on} three largely independent choices \hl{(Figure }\ref{fig:framework}\hl{)}: (1) how to measure the similarity between cases, (2) how to retrieve a cohort of cases similar to an index case, and (3) how to generate a prediction or other statistical insight for the index case from the similarity cohort.

@@

\hl{As in the case of data and, recently, of ML models, having a standardized general framework facilitates the recombination of ideas and prevents duplicate effort as well as supporting more comprehensive and efficient work [@Ravi2022]. A SCBM implementation based on this framework would help investigators systematically compare predictions and other output obtained using different tools for cohort retrieval and predictive modeling, as well as tune the parameters governing these tools. It would also make like-kind comparisons of performance easier on different tasks using different data sets and expedite validation studies. This would address the core technical needs expressed in our sample (Section }\ref{identified-needs}\hl{) as well as the limitations of our own comparison of reported performances (Section }\ref{evaluations}\hl{).
We also believe that this framework could be used to make the diversity of possible approaches more comprehensible to users, enabling them to most appropriately tailor these decisions to their tasks.}

@@

\hl{Most studies were premised on the potential predictive value of localized models (Appendix), but not all experiments affirmed this premise (Section }\ref{evaluations}\hl{). Those that did mostly used DML to improve retrieval and sometimes iteratively optimized the metric and the model.
However, even those that showcased several such specializations }[@CampilloGimenez2013; @Ng2015; @Zhang2018; @Liu2022]\hl{ mostly used their own precursors or commonplace alternatives as comparators. Future work should compare competitive alternatives to assess their performance and other trade-offs.
This work would be aided by standardization and modularization using a framework like that outlined above.
For example, }@CampilloGimenez2013\hl{ and }@Liu2022\hl{ use a modular design to toggle multiple choices and obtain neatly nested comparisons.}

@@

The search \hl{was} designed to recover studies of the kind reviewed by the review papers from which we obtained our “seed set”. To validate the final search design, we \hl{determined} how many of the papers in this seed set that are indexed by PubMed are actually recovered by our search. In most cases, the focus of a review paper is different from ours, so we will only perform this validation test on the seed set obtained from two review papers that are (a) closest in focus to ours and (b) use terminology associated with the two distinct sub-literatures relevant to our focus: Choudhury & Begum (2016), which focuses on case-based reasoning in medicine, and Sharafoddini, Dubin, & Lee (2017), which focuses on patient similarity–based prediction models on health data. The proportion of each PubMed-indexed seed set \hl{recovered} from our PubMed search provides a rough and optimistic yet useful estimate of the proportion of the relevant literature that our full search strategy will recover.

Once we \hl{finalized} the search as a logical pattern, we \hl{took} the following steps to generate the sample/corpus of literature that \hl{was} the starting point for our selection process.

1. The logical pattern will be converted to a search string using the syntax appropriate to each database in our search strategy. These include\hl{d} PubMed (already done as part of the search design), Web of Science, Academic Search Premier/Elite, and Mathematical Reviews.
2. The search \hl{was} conducted on each database and the results organized into a Zotero collection, with one subcollection for each database.
3. Duplicate results \hl{were} identified and merged. (A result obtained from multiple databases should have only one Zotero entry but should be filed under the subcollection for each database in which it was found.)

@@

These programs used varying terminology for common concepts, and no common term is in use for what we here refer to as \hl{similarity cohort--based modeling}; authors described their approaches as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020; @Liu2022; @Doborjeh2022], and "precision cohort" analysis [@Wang2019; @Tang2021; @Ng2021]. \hl{(Interestingly, none of the articles described their methods as ``explainable'' ML.)} We find uses of "targeted", "personalized", and "precision" generic and imprecise, while transductive inference is an established term for a broader set of methods in ML.
\hl{While \emph{local} naturally contrasts with \emph{global}, the overloaded term \emph{localized models} stands to cause more confusion than it resolves.}

@@

\hl{Our review was driven by an interest in the potential of SCBM toward interpretable ML. While not essential to the main text, we were therefore interested in understanding what needs motivated the various proposed methods we found.}

### 2025-03-14 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.tex
put latex revisions into rmarkdown

%\usepackage{soul}
\newcommand{\hl}[1]{#1}

@@

We conducted all searches over 2021 July 19--22. We tentatively excluded
results from Google Scholar because they were missing abstracts, and
later agreed to discard these results due to the lack of reproducibility
of the search.
We repeated the search on 2024 January 24 on the PubMed and Web of
Science databases to bring the results up to date.

@@

\subsection{\texorpdfstring{Challenges\hl{ faced}}{Challenges}}\label{challenges}

@@

\hl{As in the case of data and, recently, of ML models, having a standardized general framework facilitates the recombination of ideas and prevents duplicate effort as well as supporting more comprehensive and efficient work (Ravi et al.). A SCBM implementation based on this framework would help investigators systematically compare predictions and other output obtained using different tools for cohort retrieval and predictive modeling, as well as tune the parameters governing these tools. It would also make like-kind comparisons of performance easier on different tasks using different data sets and expedite validation studies. This would address the core technical needs expressed in our sample (Section }\ref{identified-needs}\hl{)
as well as the limitations of our own comparison of reported
performances (Section }\ref{evaluations}\hl{).
We also believe that this framework could be used to make the diversity of possible approaches more comprehensible to users, enabling them to most appropriately tailor these decisions to their tasks.}

@@

Once we \hl{finalized} the search as a logical pattern, we \hl{took} the
following steps to generate the sample/corpus of literature that will be
the starting point for our selection process.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The logical pattern \hl{was} converted to a search string using the
  syntax appropriate to each database in our search strategy. These
  include\hl{d} PubMed (already done as part of the search design), Web of
  Science, Academic Search Premier/Elite, and Mathematical Reviews.
\item
  The search \hl{was} conducted on each database and the results
  organized into a Zotero collection, with one subcollection for each
  database.
\item
  Duplicate results \hl{were} identified and merged. (A result obtained
  from multiple databases should have only one Zotero entry but should
  be filed under the subcollection for each database in which it was
  found.)
\end{enumerate}

@@

\subsection*{\hl{Terminology}}\label{terminology}
\addcontentsline{toc}{subsection}{Terminology}

@@

\subsection*{\texorpdfstring{Rationales\hl{ for proposed methods}}{Rationales}}\label{rationales}

@@

%\textbf{TODO: Place vertical separators between major columns.}

>>>>

\usepackage{soul}

@@

We conducted all searches over 2021 July 19--22. We tentatively excluded
results from Google Scholar because they were missing abstracts, and
later agreed to discard these results due to the lack of reproducibility
of the search.

We repeated the search on 2024 January 24 on the PubMed and Web of
Science databases to bring the results up to date.

@@

\subsection{\texorpdfstring{Challenges\hl{ with methods}}{Challenges}}\label{challenges}

@@

\hl{As in the case of data and, recently, of ML models, having a standardized general framework facilitates the recombination of ideas and prevents duplicate effort as well as supporting more comprehensive and efficient work [@Ravi2022]. A SCBM implementation based on this framework would help investigators systematically compare predictions and other output obtained using different tools for cohort retrieval and predictive modeling, as well as tune the parameters governing these tools. It would also make like-kind comparisons of performance easier on different tasks using different data sets and expedite validation studies. This would address the core technical needs expressed in our sample (Section }\ref{identified-needs}\hl{) as well as the limitations of our own comparison of reported performances (Section }\ref{evaluations}\hl{).
We also believe that this framework could be used to make the diversity of possible approaches more comprehensible to users, enabling them to most appropriately tailor these decisions to their tasks.}

@@

Once we \hl{finalized} the search as a logical pattern, we \hl{took} the
following steps to generate the sample/corpus of literature that
\hl{was} the starting point for our selection process.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The logical pattern will be converted to a search string using the
  syntax appropriate to each database in our search strategy. These
  include\hl{d} PubMed (already done as part of the search design), Web
  of Science, Academic Search Premier/Elite, and Mathematical Reviews.
\item
  The search \hl{was} conducted on each database and the results
  organized into a Zotero collection, with one subcollection for each
  database.
\item
  Duplicate results \hl{were} identified and merged. (A result obtained
  from multiple databases should have only one Zotero entry but should
  be filed under the subcollection for each database in which it was
  found.)
\end{enumerate}

@@

\subsection*{Terminology}\label{terminology}
\addcontentsline{toc}{subsection}{Terminology}

@@

\subsection{\texorpdfstring{Rationales\hl{ for methods}}{Rationales}}\label{rationales}

@@

\textbf{TODO: Place vertical separators between major columns.}
